Dynamical Stochastic Approximation Monte Carlo

### 2018-04-11 papers/histogram/sad.tex
include gamma plot in sad paper (first draft of plot)

\section{Introduction}
Over the past several decades a number of advanced Monte-Carlo
simulation algorithms have been developed which calculate the
properties of a thermal system over a range of temperatures.  This
idea began with the original histogram method, which used a single
canonical Monte Carlo simulation to predict properties for nearby
temperatures~\cite{ferrenberg1988new}.  For large simulations this
approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This
led to a variety of ``broad'' (or ``flat'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
  wang2001determining, wang2001efficient, trebst2004optimizing}, which
attempt to explore a wide range of energies.  These methods also
benefit in that they are unlikely to be trapped in a local energy
minimum.

Wang and Landau introduced an algorithm that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has several user-defined
parameters.  This can make its application to a variety of systems
something of an art-form since the user needs to determine the ideal
parameters for his particular system.  Also, detailed balance is
violated (although for only short periods of time), yet the method
converges in most situations.  This raises several questions such as
when does the method converges, what rate will the method converge to
the correct density of states, and how to appropriately decide what
parameters to choose so that the algorithm solves the problem in the
best way. Transition Matrix Monte Carlo~\cite{wang1999transition,
swendsen1999transition, fitzgerald2000monte} became an attractive
complimentary simulation algorithm to Wang-Landau in that, if detailed
balance is ensured, the system is guaranteed to converge (something
which is not guaranteed for the Wang-Landau Algorithm).  Unfortunately,
the algorithm can be considerably more difficult to implement when
compared to Wang-Landau: in particular, the infinite temperature
transition matrix~\cite{wang2002transition}.  Also, the time necessary
for the density of states to converge could be considerable
due to the algorithm spending too much time exploring low energy states.
~\tssays{(note to self) The above paragraph could be condensed a bit}

Shell et al.~\cite{shell2003improved, shell2004flat} originially
implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
effort to quickly explore the energies of a given system using WL then
switch over to TMMC to guarantee convergence. Considerable effort has
been spent trying to determine the flatness criteria and cutoff for the
WL portion, but there remains no rigorous systematic way to determine
these parameters~\cite{rane2013monte}.  The algorithm is first run for
several trial simulations and/or `experience' is used to determine the
best parameters for a users particular system of
study~\cite{siderius2013use}.  Around the same time, the convergence
rate for Wang-Landau was being explored~\cite{zhou2005understanding}.
It was found that utilizing modification factors that decrease faster
than $1/t^2$ leads to nonconvergence~\cite{belardinelli2007fast}.  This
lead to the so-called $1/t$-WL algorithm which ensures that excess CPU
time is not wasted by continuing to perfrom calculations once the error
in the density of states becomes
saturated~\cite{belardinelli2008analysis}. An effort to reduce the
number of user defined parameters of WL and formulate a theory for why
the method converges despite detailed balance being violated although
infrequently was also being udertaken.  Liang, Liu, and Carrol began to
consider whether WL could be considered a special case of Monte Carlo
whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form
of Stochastic Approximation Monte Carlo (SAMC).  While SAMC can
guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

In our work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamical
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several histogram methods including TMMC,
WL, WL-TMMC, and SAMC.

@@

Here we employ a variety of broad-histogram methods.  We outline the
general workings of algorithm we developed in detail and summarize
algorithms that can be found elsewhere.  The following algorithms are
introduced and applied to the square-well fluid:  Transition
Matrix Monte-Carlo (TMMC), Wang-Landau (WL), Wang-Landau Transition
Matrix Monte-Carlo (WL-TMMC), Stochastic Approximation Monte-Carlo
Monte-Carlo (SAMC), and Dynamical Stochastic-Approximation (SAD).

@@

\subsubsection{define a transition probability}
Assuming that the system under study is in a state denoted old and a
proposed transition is made to a new state denoted new, the probability
is defined as
\begin{align}
  p_{old \rightarrow new} = p_{new \rightarrow old}
\end{align}
where we have allowed the probabilies to be equal for simplicity of presentation.

\subsubsection{define an acceptance probability}
The probability for accepting a transition from an old state to a new one can be
defined
\begin{align}
  p_{a} = \min\bigg[1,\frac{\alpha_{new\rightarrow old}}
  {\alpha_{old \rightarrow new}}\frac{\pi_{new}}{\pi_{old}}\bigg]
\end{align}
where $\alpha_{new\rightarrow old}$ is the probability of generating a
new system configuration the old.  For conventional Monte-Carlo moves
$\alpha_{old \rightarrow new} =\alpha_{new\rightarrow old}$ although
for advanced Monte-Carlo moves this would not be the
case~\cite{paluch2008comparing, siepmann1990method}.  The probabily of
observing the system in the old state can be found based on the
ensemble constraints placed on the system.  Fitzgerald et al.
considered the system of interest using the canonical ensemble however
extension to other ensembles is quite natural.
\begin{align}
  \pi_{old} = \frac{e^{-\beta E_{old}}}{Z}
\end{align}
The acceptance probability given for TMMC is idential to that introduced by
Swendsen and Wang~\cite{wang2002transition}.
\begin{align}
  p_{a} = \min\bigg[1,e^{-\frac{\delta\epsilon}{k_{B} T}}\bigg]
\end{align}
\subsubsection{define a bookkeeping step}
The third step in the TMMC algorithm consists of recording the data in a collection
matrix (analogous to the transition histogram matrix).
\begin{align}
  C_{N,\delta} = C_{N,\delta} + p_{a}\\
  C_{N,0} = C_{N,0} +(1 - p_{a})
\end{align} If $\delta=0$, that is, the energy state has not been
changed, then only $C_{N,0}$ is incremented by unity. Fitzgerald et
al.~\cite{fitzgerald2000monte} show that defining the third step this
way results in a uniform improvement over using histogram estimators to
determine the particle number probability distribution (PNPD).

@@

The Wang-Landau Algorithm developed by Wang and Landau proposes a
random walk in energy space in order to estimate the density of
states~\cite{wang2001efficient}. While methods such as Metropolis
Sampling and Swendsen-Wang cluster
flipping~\cite{swendsen1987nonuniversal} generate a narrow distribution
and require sampling at individual temperatures, the Wang-Landau
algorithm uses a flat histogram and performs a random walk in energy
space to determine the density of states~\cite{wang2001determining,
landau2014guide}. Wang-Landau's major tenant is that when counting the
histogram $H(\epsilon)$, the energy occurences should form a flat
distribution.  The criteria for flat sampling is given by
\begin{equation}
	\frac{\min_{\epsilon} H(\epsilon)}
	{\big\langle H(\epsilon)\big\rangle }
	> \eta
\end{equation} where $\eta$ is usually set to $0.80$ and determines how
many times each energy is sampled relative to the mean number of
visits.  Difficulties can occur with this flatness criteria due to the
fact that some energies on a given energy range might never be
sampled~\cite{haber2014transition}.  Over the past decade much
attention was directed at examining the rate of convergence and error
saturation~\cite{lee2006convergence, belardinelli2007wang}.  The
algorithm is briefly outlined below where $w$ represents the weights:
\begin{equation}
	\ln{w_{k+1}(\epsilon,N)}=\ln{w_{k}(\epsilon,N)}
	+\gamma
\end{equation}
\begin{equation}
	\ln{\gamma_{k+1}}=\frac{u}{2}\ln{\gamma_{k}}
\end{equation}
for $u$ typically greater than 1 and $\gamma$ is the modification factor for
the density of states.  We then accept a new configuration for the density of
states with a probability given by the following
\begin{equation}
	\mathcal{P}(\epsilon_i \rightarrow \epsilon_j)
	= \min[1,e^{\ln{w(\epsilon_i)}-\ln{w(\epsilon_j)}}]
\end{equation}
Zhou and Bhatt explored the convergence of the Wang-Landau algorithm and
determined that choosing a large modification factor $\gamma_0 = e^{4}$ and
rapidly reducing the factor by 10 during each step resulted in reduced statistical
error~\cite{zhou2005understanding}.

@@

\paragraph{Initializing Wang-Landau:} We begin the simulation by
defining a minimum important energy and a maximum entropy state. In
traditional Wang-Landau the WL factor begins at 1.0 and the factor is
modified by 2.0 until the simulation is cutoff typically around
$10^{-10}$.  Rane et. al. published an implemenation of
WL-TMMC~\cite{rane2013monte} that suggested using a WL factor between 1.0 and
$10^{-2}$ with a cutoff $<10^{-5}$. In our implementation of WL-TMMC,
we set the WL factor to be 1 and the cutoff at $10^{-10}$. A single
sweep is defined as each macrostate being sampled a number of times in
this case 10,000 (Siderius et. al. does 100). Wang-Landau sets a
flatness criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually between 0.75 to 0.90.  For our simulation, it is
sufficient that each macrostate is visited at least once (each
histogram bin has one entry)~\cite{shell2003improved}.  The goal of
this WL initialization is to prefill the transition matrix for the next
portion of the simulation.
\paragraph{Initializing TMMC:} The purpose of prefilling the transition
matrix is apparent because if numerous zeros existed in the collection
matrix, the infinite temperature transition matrix would be
ill-defined.  Sampling over large ranges of density of states would
therefore take an unacceptable length of time to
complete~\cite{shell2003improved, shen2014elucidating}.  The transition
matrix in terms of the collection matrix is given as follows:
\begin{align}
\widetilde{T}_{\infty}(I\rightarrow J) = \frac{C(I,J)}
{\sum_{K} C(I,K)}
\end{align}
We now run the TMMC portion of the simulation.  Siderius does this for
20 additional sweeps (does about 8 sweeps for WL).  We do this until we
reach a specified temperature of interest.
We have not yet implemented a biasing function to control acceptance
rate of trial moves but we probably could/should? at some point.
\begin{align}
  P_{acc}(i\rightarrow j) = \min\bigg[1,\frac{\widetilde{T}_{\infty}(J\rightarrow I)}
  {\widetilde{T}_{\infty}(I\rightarrow J)}\bigg]
\end{align}
this is only used to control actual transitions not update the collection
matrices.

\subsection{SAD Algorithm}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. SAMC can be
applied as a dynamic importance sampling algorithm that uses an update
factor to explore energy space~\cite{liang2007stochastic,
werlich2015stochastic, schneider2017convergence}.  The update factor is
defined in the original implementation~\cite{liang2007stochastic} in
terms of an arbitrary tunable parameter $t_0$.
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
We argue that it is possible to set this parameter to a system
independent parameter such that $t_0 = N_E^2/t$ where $N_E$ is the
number of energies that have been found.  One of the energy found terms
can be thought of as a change in the entropy $\Delta S$.  The change in
the entropy can be thought of as a slope of the triangle made by
connecting the minimum important energy and the maximum entropy state
divided by the minimum temperature:
\begin{align}
\Delta S = \frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
This gives an update factor as follows:
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{N_E\Delta S}{t} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{T_{\text{min}}}
\end{align}
The update factor has the same $1/t$ behavior as the original SAMC
algorithm; however now every time a new energy is discovered during the
course of the simulation, the update factor experiences a jump.
~\jpsays{Add in additional material for SAD!}

>>>>

\section{Introduction}
Over the past several decades a number of broad histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for all temperatures.  The
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large simulations
this approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This led
to a variety of ``broad'' (or ``flat'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
wang2001determining, wang2001efficient, trebst2004optimizing}, which
attempt to explore a wide range of energies.  These methods also
benefit in that they are unlikely to be trapped in a low enetropy state.

Wang and Landau introduced an algorithm that uses an update factor and
a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has several user-defined parameters.  This
can make its application to a variety of systems something of an
art-form since the user needs to determine the ideal parameters for the
particular system being studied.  Also, detailed balance is violated
(although only for short periods of time), ensuring that convergence is
not guaranteed.  This raises several questions such as when the method
does converge, what rate will the method converge to the correct
density of states. Also, how can the user appropriately decide what
parameters to choose so that the algorithm solves the problem in the
most ideal way.

Transition Matrix Monte Carlo~\cite{wang1999transition,
swendsen1999transition, fitzgerald2000monte} became an attractive
complimentary simulation algorithm to Wang-Landau, in that, if detailed
balance is ensured, the system is guaranteed to converge.  Unfortunately,
the algorithm can be considerably more difficult to implement when
compared to Wang-Landau: in particular, due to the infinite temperature
transition matrix~\cite{wang2002transition}.  Also, the time necessary
for the density of states to converge could be considerable
due to the algorithm spending too much time exploring low energy states.
Unlike WL, TMMC does not require a prior knowledge of the energy range of
the system.

Shell et al.~\cite{shell2003improved, shell2004flat} originially
implemented Wang-Landau Transition Matrix Monte Carlo (WL-TMMC) in an
effort to quickly explore the energies of a given system using WL then
switch over to TMMC to guarantee convergence. Considerable effort has
been spent trying to determine the flatness criteria and cutoff for the
WL portion, but there remains no rigorous systematic way to determine
these parameters~\cite{rane2013monte}.  The algorithm is first run for
several trial simulations and/or `experience' is used to determine the
best parameters for a users particular system of
study~\cite{siderius2013use}.

Around the same time, the convergence rate for WL was being
explored~\cite{zhou2005understanding,lee2006convergence,
belardinelli2007wang}. It was found that utilizing modification factors
that decrease faster than $1/t^2$ leads to
nonconvergence~\cite{belardinelli2007fast}.  This lead to the so-called
$1/t$-WL algorithm which ensures that excess CPU time is not wasted by
continuing to perfrom calculations once the error in the density of
states becomes saturated~\cite{belardinelli2008analysis}. An effort to
reduce the number of user defined parameters of WL and formulate a
theory for why the method converges despite detailed balance being
violated although infrequently was also being udertaken.

Liang, Liu, and Carrol began to consider whether WL could be considered
a special case of Stochastic Approximation whose convergence could be
mathematically proven~\cite{liang2006theory, liang2007stochastic}. In
2007, Liang et al.~\cite{liang2007stochastic} argued that WL can be
considered a form of Stochastic Approximation Monte Carlo (SAMC). While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

In this work, we have developed a novel algorithm based on SAMC that
does not require user-defined inputs and therefore should be easily
applicable to a given system.  We call this method SAD (Dynamical
Stochastic Approximation), and will discuss it in detail in the methods
section. We compare it along with several broad histogram methods
which include TMMC, WL, WL-TMMC, and SAMC.

@@

In this work, We compare a variety of broad-histogram methods.  We
outline the general workings of each algorithm that we developed in
detail while summarizing algorithms that can be found elsewhere.  The
following methods are introduced and applied to the square-well fluid:
Transition Matrix Monte-Carlo (TMMC), Wang-Landau (WL), Wang-Landau
Transition Matrix Monte-Carlo (WL-TMMC), Stochastic Approximation
Monte-Carlo (SAMC), and Dynamical Stochastic-Approximation
(SAD).

@@

\subsubsection{define a transition probability}
Assuming that the system under study is in a state denoted old, and a
proposed transition is made to a new state denoted new, the probability
is defined as
\begin{align}
  p_{old \rightarrow new} = p_{new \rightarrow old}
\end{align}
where we have allowed the probabilies to be equal for simplicity of
presentation.

\subsubsection{define an acceptance probability}
The probability for accepting a transition from an old state to a new
one can be defined
\begin{align}
  p_{a} = \min\bigg[1,\frac{\alpha_{new\rightarrow old}}
  {\alpha_{old \rightarrow new}}\frac{\pi_{new}}{\pi_{old}}\bigg]
\end{align}
where $\alpha_{new\rightarrow old}$ is the probability of generating a
new system configuration the old.  For conventional Monte-Carlo moves
$\alpha_{old \rightarrow new} =\alpha_{new\rightarrow old}$ although
for advanced Monte-Carlo moves this would not be the
case~\cite{paluch2008comparing, siepmann1990method}.  The probabily of
observing the system in the old state can be found based on the
ensemble constraints placed on the system.  Fitzgerald et al.
considered the system of interest using the canonical ensemble;
however, the extension to other ensembles is quite natural.
\begin{align}
  \pi_{old} = \frac{e^{-\beta E_{old}}}{Z}
\end{align}
The acceptance probability given for TMMC is idential to that
introduced by Swendsen and Wang~\cite{wang2002transition}.
\begin{align}
  p_{a} = \min\bigg[1,e^{-\frac{\delta\epsilon}{k_{B} T}}\bigg]
\end{align}
\subsubsection{define a bookkeeping step}
The third step in the TMMC algorithm consists of recording the data in
a collection matrix (the unormalized transition histogram matrix).
\begin{align}
  C_{N,\delta} = C_{N,\delta} + p_{a}\\
  C_{N,0} = C_{N,0} +(1 - p_{a})
\end{align} If $\delta=0$, that is, the energy state has not been
changed, then only $C_{N,0}$ is incremented by unity. Fitzgerald et
al.~\cite{fitzgerald2000monte} showed that defining the third step this
way results in a uniform improvement over using histogram estimators to
determine the particle number probability distribution (PNPD).

@@

The Wang-Landau Algorithm developed by Wang and Landau proposes a
random walk in energy space in order to estimate the density of
states~\cite{wang2001efficient,wang2001determining, landau2014guide}.
Wang-Landau's major tenant is that when counting the histogram
$H(\epsilon)$, the energy occurences should form a flat distribution.
The criteria for flat sampling is given by
\begin{equation}
	\frac{\min_{\epsilon} H(\epsilon)}
	{\big\langle H(\epsilon)\big\rangle }
	> \eta
\end{equation}
where $\eta$ is usually set to $0.80$ and determines how many times
each energy is sampled relative to the mean number of visits.
Difficulties can occur with this flatness criteria due to the fact that
some energies on a given energy range might never be
sampled~\cite{haber2014transition}. The algorithm is briefly outlined
below where $w$ represents the weights:
\begin{equation}
	\ln{D_{i+1}(\epsilon,N)}=\ln{D_{i}(\epsilon,N)}
	+\gamma
\end{equation}
\begin{equation}
	\ln{\gamma_{i+1}}=\frac{u}{2}\ln{\gamma_{i}}
\end{equation}
for $u$ typically greater than 1 and $\gamma$ is the modification
factor for the density of states.  A new configuration for the density
of states with a probability given by the following is chosen:
\begin{equation}
	\mathcal{P}(\epsilon_\text{old} \rightarrow \epsilon_\text{new})
	= \min[1,e^{\ln{D(\epsilon_\text{old})}-\ln{D(\epsilon_\text{new})}}]
\end{equation}
The entire process is repeated until a desired cutoff is reached.

@@

\paragraph{Initializing Wang-Landau:} We begin the simulation by
defining a minimum important energy and a maximum entropy state. In
traditional Wang-Landau the WL factor begins at 1.0 and the factor is
modified by 2.0 until the simulation is cutoff typically around
$10^{-10}$.  Rane et. al. published an implemenation of
WL-TMMC~\cite{rane2013monte} that suggested using a WL factor between
1.0 and $10^{-2}$ with a cutoff $<10^{-5}$. In our implementation of
WL-TMMC, we set the WL factor to be 1 and the cutoff to be either
$10^{-4}$ or $10^{-10}$. A single sweep is defined as each macrostate
being sampled a number of times in this case 10,000 (although 100 is
often sufficient for small systems). Wang-Landau sets a flatness
criteria~\cite{wang2001determining, wang2001efficient,
hatch2015computational, mahynski2017predicting} for the accumulated
histogram usually between 0.75 to 0.90.  In the original
implementation, it was found to be sufficient that each macrostate is
visited at least once (each histogram bin has one
entry)~\cite{shell2003improved}; however, for other systems this is
rarely sufficient.  The goal of this WL initialization is to prefill
the transition matrix for the next portion of the simulation.
\paragraph{Initializing TMMC:} The purpose of prefilling the transition
matrix is apparent because if numerous zeros existed in the collection
matrix, the infinite temperature transition matrix would be
ill-defined.  Sampling over large ranges of density of states would
therefore take an unacceptable length of time to
complete~\cite{shell2003improved, shen2014elucidating}.  The transition
matrix in terms of the collection matrix is given as
follows:
\begin{align}
\widetilde{T}_{\infty}(I\rightarrow J) = \frac{C(I,J)}
{\sum_{K} C(I,K)}
\end{align}
We now run the TMMC portion of the simulation.
%Siderius does this for 20 additional sweeps (does about 8 sweeps for
%WL).  We do this until we reach a specified temperature of interest. We
%have not yet implemented a biasing function to control acceptance rate
%of trial moves but we probably could/should? at some point.
The acceptance probability is written in terms of the infinite temperature
transition matrix.
\begin{align}
  P_{acc}(i\rightarrow j) = \min\bigg[1,\frac{\widetilde{T}_{\infty}(J\rightarrow I)}
  {\widetilde{T}_{\infty}(I\rightarrow J)}\bigg]
\end{align}
This probability is only used to control actual transitions not update
the collection matrices.

\subsection{SAMC Algorithm}
SAMC can be applied as a dynamic importance sampling algorithm that
uses an update factor to explore energy
space~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$ and a scale factor $\gamma_0$.
\begin{align}
\gamma_{t}^{\text{SA}} =\gamma_0 \frac{t_0}{\max(t_0,t)}
\end{align}
SAMC offers several advantages over other broad histogram methods
including proven convergence. It has been shown that if
$F_{t}^{\text{SA}}$ satisfies two conditions the estimator is proven to
converge to the real DOS~\cite{liang2006theory, liang2007stochastic}.
\begin{align}
\sum_{t=1}^\infty \gamma_{t}^{\text{SA}} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty (\gamma_{t}^{\text{SA}})^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$.  Unlike WL and WL-TMMC, the energy range need
not be known a priori.  The time to converge depends only on the choice of
parameters $t_0$ and $\gamma_0$.
Unfortunately, only by performing many computational runs might the user
determine what SAMC arbitrary parameters to choose.  This has lead us
to develop a systematic way (which we will discuss in the next section)
for the simulation to determine what $t_0$ should be.

\section{SAD Algorithm}
\subsubsection{The theory behind SAD}
The SAD (Dynamic Stochastic Approximation) is a version of the SAMC
Algorithm that attempts to dynamically choose the modification factor
rather than relying on system dependent parameters such as $t_0$ or
$\gamma_0$.  There is an immediate advantage of such an algorithm where
parameters are chosen independent of system size or type. While for
SAMC the update factor is defined in the original
implementation~\cite{liang2007stochastic} in terms of an arbitrary
tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} = \frac{t_0}{\max(t_0,t)}
\end{align}
The update factor $\gamma_{t}^{\text{SA}}$ can be thought of as
$\nicefrac{\text{d}S}{\text{d}t}$. This tells us that $t_0$ should have
dimensions of entropy.  We argue that we can write this entropy `term'
geometrically as an area bounded by the $\ln{D(E)}$ i.e. the entropy.  This
gives the following
\begin{align}
\gamma_{t}^{\text{SA}} =
\frac{N_E}{t}\frac{E_{\text{max}}-E_{\text{min}}}{3T_{\text{min}}}
\end{align}
where $N_E$ is the number of energies that have been found. The change
in the entropy the minimum temperature can be thought of as a slope of
the triangle made by connecting the minimum important energy and the
maximum entropy state .  The factor of $3$ comes from the area bounded
by the $\ln{D(E)}$ The update factor has the same $1/t$ behavior as the
original SAMC algorithm; however now every time a new energy is
discovered during the course of the simulation, the
update factor experiences a jump.

\begin{figure}
  \includegraphics[width=\columnwidth]{figs/gamma-ww1_30-ff0_22-100x10}
  \caption{The update factor $\gamma$ versus iteration for three
    different methods: WL, SAMC, and SAD.
    method.}\label{fig:gamma-vs-t}
\end{figure}

Fig.~\ref{fig:gamma-vs-t}. shows the monotonic decrease of the update
factor for SAMC.  The update factor for WL decreases like a step
function while the update factor for SAD dynamically fluctuates around
a value less than 1 for early MC moves (while finding new energy
states).  Ultimately once all energy states have been found, the update
factor for SAD decreases in the same monotonic fashion as SAMC.

\subsubsection{Energy binning with SAD}

Energy binning is an important consideration for complex continous
systems. Coarse binning yields fewer energy states for the walker to
explore but more time spent in the individual bin. Fine binning yields
more energy states to explore but less time spent in each bin.  Some
methods such as AdaWL~\cite{koh2013dynamically} (discussed elsewhere)
employ a tunable mechanism for controlling the binning for low entropic
states in order to ensure the exploration of all energies. A
significant advantage of SAD over other broad histogram methods is that
energy binning is solved dynamically. SAD should perform independently
for a reasonable range of binning because $\gamma \propto
\nicefrac{N_{E}}{t}$.  As the number of energy states found $N_{E}$
increases (fine binning), the time spent $t$ in each bin will decrease.
This argument explains why SAD should perform well for complex systems
in terms of binning. SAMC has a prefactor $\gamma_0$ to aid in a
similar way but this adds yet another parameter for the user to choose.

\subsubsection{Scaling, step size, and SAD}
Methods such as WL can fail to converge for complex systems in part
because the same random MC moves are performed for the entire energy
range.  The main problem lies in trying to determine how to optimize
the step size such that the time to explore all energies is reasonable.
Small step sizes lead to improved sampling at the expense of slow
convergence. Koh et al. proposed multiplying $\gamma$ by an energy
dependent factor in order to improve the converge of WL in regards to
convergence and scaling~\cite{koh2013dynamically}.  Because the update
factor for SAD has an energy dependent factor already present (and
dynamically adjusted), we expect that SAD will perform consitently
scaling free for a reasonable choice of step size.  SAD should perform
extremely well compared with other broad histogram methods when applied
to complex continuous systems such as polymers/proteins and spin models.
